Structured Versus Decorated Cospans

### 2022-04-17 cospans_compositionality_4.tex
I will try to fix the Petri net problems in this version
I tried to fix section 6.3.  Check it out!

% Structured versus decorated cospans
% John Baez, Kenny Courser and Christina Vasilakopoulou
% in LaTeX
% 2021/04/08 - third version sent to Compositionality: version for publication

@@

We thank Daniel Cicala, Brendan Fong, Joe Moeller and Morgan Rogers for helpful conversations, and the referees for their many useful suggestions.  We especially thank Michael Shulman for suggesting a proof strategy that greatly improved this paper.   The third author would like to thank the General Secretariat for Research  and Technology (GSRT) and the Hellenic Foundation for Research and Innovation (HFRI).

@@

Pollard and the first author showed how to construct an open dynamical system from any open Petri net with rates,  thus defining a functor from a category with open Petri nets with rates as morphisms to one with open dynamical systems as morphisms \cite{BP}.  They used Fong's original decorated cospans to do this.   Later, structured cospans were used to promote the first of these categories to a double category \cite[Section 6.16]{BC}. Here we show that the second of these categories, with open dynamical systems as morphisms, \emph{cannot} be made into a double category using structured cospans.  However, we \emph{can} do it using decorated cospans.

@@

Now we turn to the details.  A \define{Petri net with rates} is a Petri net $s,t \maps T \to \N[S]$ together with a function $r \maps T \to (0,\infty)$ assigning to each transition $\tau \in T$ a positive real number called its \define{rate constant}.  There is a category $\Petri_r$ whose objects are Petri nets with rates, where a morphism from 
\[   \xymatrix{ (0,\infty) & T \ar[l]_-r \ar@<-.5ex>[r]_-t \ar@<.5ex>[r]^-s & \N[S] }\]
 to 
 \[   \xymatrix{ (0,\infty) & T' \ar[l]_-{r'} \ar@<-.5ex>[r]_-{t'} \ar@<.5ex>[r]^-{s'} & \N[S'] }\]
is a morphism of the underlying Petri nets such that the following diagram also commutes:
\[
\begin{tikzpicture}[scale=1.5]
\node (E) at (3,-0.5) {$(0,\infty)$};
\node (G) at (4,0) {$T$};
\node (G') at (4,-1) {$T'$};
\path[->,font=\scriptsize,>=angle 90]
(G) edge node[right]{$g$} (G')
(G) edge node[above]{$r$} (E)
(G') edge node[below]{$r'$} (E);
\end{tikzpicture}
\]
There is a functor $R \maps \Petri_r \to \Set$ sending any Petri net with rates to its set of places, and this has a left adjoint $L \maps \Set \to \Petri_r$ sending any set $S$ to the Petri net with $S$ as its set of places and no transitions \cite[Lemma 6.18]{BC}.   Since $\Petri_r$ has finite colimits \cite[Lemma 6.19]{BC}, it follows that there is a symmetric monoidal double category ${}_L \lCsp(\Petri_r)$ where:
\begin{itemize}
\item objects are finite sets,
\item vertical 1-morphisms are functions,
\item horizontal 1-cells are \define{open Petri nets with rates}, namely diagrams in $\Petri_r$ of the form
\[
\begin{tikzpicture}[scale=1.5]
\node (D) at (-3,0) {$L(X)$};
\node (E) at (-2,0) {$P$};
\node (F) at (-1,0) {$L(Y)$,};
\path[->,font=\scriptsize,>=angle 90]
(D) edge node [above] {$i$} (E)
(F) edge node [above] {$o$} (E);
\end{tikzpicture}
\]
\item 2-morphisms are diagrams in $\Petri_r$ of the form
\[
\begin{tikzpicture}[scale=1.5]
\node (E) at (3,0) {$L(X)$};
\node (F) at (5,0) {$L(Y)$};
\node (G) at (4,0) {$P$};
\node (E') at (3,-1) {$L(X')$};
\node (F') at (5,-1) {$L(Y')$.};
\node (G') at (4,-1) {$P'$};
\path[->,font=\scriptsize,>=angle 90]
(F) edge node[above]{$i$} (G)
(E) edge node[left]{$L(f)$} (E')
(F) edge node[right]{$L(g)$} (F')
(G) edge node[left]{$\alpha$} (G')
(E) edge node[above]{$o$} (G)
(E') edge node[below]{$i'$} (G')
(F') edge node[below]{$o'$} (G');
\end{tikzpicture}
\]
\end{itemize}

We can equivalently describe open Petri nets with rates using decorated cospans.  There is a symmetric lax monoidal pseudofunctor $F \maps (\Fin\Set, +) \to (\Cat, \times)$ such that for any finite set $S$, the category $F(S)$ has:
\begin{itemize}
\item objects given by Petri nets with rates whose set of places is $S$,
\item morphisms given by morphisms of Petri nets with rates that are the identity on the set of places.
\end{itemize}
This gives a symmetric monoidal double category $F \lCsp$, and using \cref{thm:equiv} we can show that this is isomorphic, as a symmetric monoidal double category, to ${}_L \lCsp(\Petri_r)$.

All this so far is very similar to the previous examples.  More interesting is the symmetric monoidal double category of open dynamical systems.   A dynamical system is a vector field from the perspective of a system of first-order ordinary differential equations. A Petri net with rates gives a special sort of dynamical system: an \define{algebraic} vector field on $\R^n$, meaning one where the components of the vector field are polynomials in the coordinates. We shall think of such a vector field as a special sort of function $v \maps \R^n \to \R^n$.

@@

Next, we can define a symmetric monoidal double functor
\[      \graysquare \maps F \lCsp \to D \lCsp \]
sending any open Petri net with rates to its corresponding open dynamical system.  This was already defined at the level of categories by Pollard and the first author \cite[Section 7]{BP}, who called it `gray-boxing'.   To boost this result to the double category level we use \cref{thm:functoriality}, taking the square in that theorem to be
\[
\begin{tikzpicture}[scale=1.5]
\node (A) at (0,0) {$\Fin\Set$};
\node (B) at (1.5,0) {$\Cat$};
\node (C) at (0,-1) {$\Fin\Set$};
\node (D) at (1.5,-1) {$\Cat$.};
\node (E) at (0.75,-0.5) {$\Downarrow \theta$};
\path[->,font=\scriptsize,>=angle 90]
(A) edge node[above]{$F$} (B)
(A) edge node[left]{$1$} (C)
(B) edge node[right]{$1$} (D)
(C) edge node[below]{$D$} (D);
\end{tikzpicture}
\] 
Here $\theta$ is a monoidal natural transformation given as follows.  For any finite set $S$, $\theta_S \maps F(S) \to D(S)$ maps any Petri net with rates 
\[   \xymatrix{ (0,\infty) & T \ar[l]_-r \ar@<-.5ex>[r]_-t \ar@<.5ex>[r]^-s & \N[S] }\]
to an algebraic vector field on $\R^S$, say $v$.   This vector field is defined using a standard prescription taken from chemistry, called `the law of mass action'.   Namely, for any $c \in \R^s$, we set
\[  
v(c) = \sum_{\tau \in T} r(\tau) \, ( t(\tau) - s(\tau) ) c^{s(\tau)} 
\]
where 
\[     c^{s(\tau)} = \prod_{i \in S} {c_i}^{s(\tau)_i}  \]
and we think of $t(\tau), s(\tau) \in \N[S]$ as vectors in $\R^S$.   This formula is explained in the paper with Pollard, where it is also shown that $\theta$ defines a monoidal natural transformation between functors to $(\Set,\times)$ \cite[Theorem 18]{BP}.   As such, it automatically becomes a monoidal natural transformation between the functors $F, D \maps (\Fin\Set,+) \to (\Cat,\times)$.  Thus, it defines a symmetric monoidal double functor $\graysquare \maps F \lCsp \to D \lCsp$.

>>>>

% Structured versus decorated cospans
% John Baez, Kenny Courser and Christina Vasilakopoulou
% in LaTeX
% 2021/04/16 - 4th version sent to Compositionality: version for publication - fixing mistakes

@@

We thank Daniel Cicala, Brendan Fong, Sophie Libkind, Joe Moeller and Morgan Rogers for helpful conversations, and the referees for their many useful suggestions.  We especially thank Michael Shulman for suggesting a proof strategy that greatly improved this paper.   The third author would like to thank the General Secretariat for Research  and Technology (GSRT) and the Hellenic Foundation for Research and Innovation (HFRI).

@@

Pollard and the first author showed how to construct an open dynamical system from any open Petri net with rates,  thus defining a functor from a category with open Petri nets with rates as morphisms to one with open dynamical systems as morphisms \cite{BP}.  They used Fong's original decorated cospans to do this.  Here we show we show how to promote these categories to double categories using our new approach to decorated cospans.

@@

Now we turn to the details.  A \define{Petri net with rates} is a Petri net $s,t \maps T \to \N[S]$ together with a function $r \maps T \to [0,\infty)$ assigning to each transition $\tau \in T$ a nonnegative real number called its \define{rate constant}.  There is a category $\Petri_r$ whose objects are Petri nets with rates, where a morphism from 
\[   \xymatrix{ (0,\infty) & T \ar[l]_-r \ar@<-.5ex>[r]_-t \ar@<.5ex>[r]^-s & \N[S] }\]
 to 
 \[   \xymatrix{ (0,\infty) & T' \ar[l]_-{r'} \ar@<-.5ex>[r]_-{t'} \ar@<.5ex>[r]^-{s'} & \N[S'] }\]
is a morphism of the underlying Petri nets whose map $g \maps T \to T'$ obeys
\[    r'(\tau') = \sum_{\{\tau \in T: g(\tau) = \tau'\}} r(\tau) \]
for all $\tau' \in T'$.   This definition was suggested by Sophie Libkind; it agrees with the earlier definition \cite{BP} in the case of isomorphisms, but not in general, and the difference is important here.

We can describe open Petri nets with rates using decorated cospans.  There is a symmetric lax monoidal pseudofunctor $F \maps (\Fin\Set, +) \to (\Cat, \times)$ such that for any finite set $S$, the category $F(S)$ has:
\begin{itemize}
\item objects given by Petri nets with rates whose set of places is $S$,
\item morphisms given by morphisms of Petri nets with rates that are the identity on the set of places.
\end{itemize}
This gives a symmetric monoidal double category $F \lCsp$ where the horizontal 1-cells are called \define{open Petri nets with rates}.

There is also a symmetric monoidal double category of open dynamical systems.   A dynamical system is a vector field as seen from the perspective of a system of first-order ordinary differential equations.  A Petri net with rates gives a special sort of dynamical system: an \define{algebraic} vector field on $\R^S$ for some finite set $S$, meaning a vector field whose components are polynomials in the coordinates. We shall think of such a vector field as a special sort of function $v \maps \R^S \to \R^S$.

@@

Next, we can define a symmetric monoidal double functor
\[      \graysquare \maps F \lCsp \to D \lCsp \]
sending any open Petri net with rates to its corresponding open dynamical system.  This was already defined at the level of categories by Pollard and the first author \cite[Section 7]{BP}, who called it `gray-boxing'.   To boost this result to the double category level we use \cref{thm:functoriality}, taking the square in that theorem to be
\[
\begin{tikzpicture}[scale=1.5]
\node (A) at (0,0) {$\Fin\Set$};
\node (B) at (1.5,0) {$\Cat$};
\node (C) at (0,-1) {$\Fin\Set$};
\node (D) at (1.5,-1) {$\Cat$.};
\node (E) at (0.75,-0.5) {$\Downarrow \theta$};
\path[->,font=\scriptsize,>=angle 90]
(A) edge node[above]{$F$} (B)
(A) edge node[left]{$1$} (C)
(B) edge node[right]{$1$} (D)
(C) edge node[below]{$D$} (D);
\end{tikzpicture}
\] 
Here $\theta$ is given as follows.  For any finite set $S$, $\theta_S \maps F(S) \to D(S)$ maps any Petri net with rates 
\[   \xymatrix{ (0,\infty) & T \ar[l]_-r \ar@<-.5ex>[r]_-t \ar@<.5ex>[r]^-s & \N[S] }\]
to an algebraic vector field on $\R^S$, say $v$.   This vector field is defined using a standard prescription taken from chemistry, called `the law of mass action'.   Namely, for any $c \in \R^s$, we set
\[  
v(c) = \sum_{\tau \in T} r(\tau) \, ( t(\tau) - s(\tau) ) c^{s(\tau)} 
\]
where 
\[     c^{s(\tau)} = \prod_{i \in S} {c_i}^{s(\tau)_i}  \]
and we think of $t(\tau), s(\tau) \in \N[S]$ as vectors in $\R^S$.   This formula is explained in the paper with Pollard \cite{BP}.   One can check that $\theta$ extends to a monoidal natural transformation between the functors $F, D \maps (\Fin\Set,+) \to (\Cat,\times)$.  Thus, it defines a symmetric monoidal double functor $\graysquare \maps F \lCsp \to D \lCsp$.
